---
title: "neuRoDev"
author: "Erik Bot"
date: "2025-05-25"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{neuRoDev}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r style, echo = FALSE, results = 'asis'}
library(BiocStyle)
```

```{r, echo = FALSE}
library(knitr)
```

neuRoDev is an R package that provides analysis tools to make use of the 
integrated transcriptomic references of cortex development, described in 
[Zonca&, Bot&, and Davila-Velderrain, 2025].

All the datasets needed as input are stored in the neuRoDevDB package.

All the main functionalities of neuRoDev are extensively described in a [Tutorial], 
at the following link: https://github.com/davilavelderrainlab/neuRoDev

neuRoDev contains functions to perform:

- Reference networks exploration ([Network exploration])
- Expression trends visualization across cell type and time ([Analysis tools])
- Single-cell RNAseq datasets mapping ([Mapping scRNAseq])
- Bulk RNAseq datasets mapping ([Mapping bulkRNAseq])

If you use `neuRoDev` or `neuRoDevDB` in your work, please cite

> **Developmental single-cell transcriptomic networks dissect neuronal and glial genesis and maturation in the human cortex**
>
> Asia Zonca&, Erik Bot&, and José Davila-Velderrain
>
> _Journal_ Date. doi: [doi](https://github.com/davilavelderrainlab/neuRoDev).

[Zonca&, Bot&, and Davila-Velderrain, 2025]: https://github.com/davilavelderrainlab/neuRoDev
[neuRoDevDB]: https://github.com/davilavelderrainlab/neuRoDevDB
[Network exploration]: https://github.com/davilavelderrainlab/neuRoDev
[Analysis tools]: https://github.com/davilavelderrainlab/neuRoDev
[Gene set networks]: https://github.com/davilavelderrainlab/neuRoDev
[Mapping scRNAseq]: https://github.com/davilavelderrainlab/neuRoDev
[Mapping bulkRNAseq]: https://github.com/davilavelderrainlab/neuRoDev
[Tutorial]: https://github.com/davilavelderrainlab/neuRoDev
[article]: https://github.com/davilavelderrainlab/neuRoDev
[issues]: https://github.com/davilavelderrainlab/neuRoDev/issues

```{r, echo=FALSE, message=FALSE, eval=TRUE}
sessionInfo()
```

